---
title: "Time_Analysis"
output: html_document
date: "2025-07-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Libraries

```{r setup, message=FALSE, warning=FALSE}
install.packages("lme4")
library(lme4)
install.packages("dplyr")
library(dplyr)
install.packages("tidyverse")
library(tidyverse)
install.packages("readxl")
library(readxl)
install.packages("lubridate")
library(lubridate)
install.packages("hms")
library(hms)
install.packages("stringr")
library(stringr)
install.packages("ggplot2")
library(ggplot2)
install.packages("emmeans")
library(emmeans)
```

# Load Data

```{r}
# Adjust this path to your file location
data <- read_excel("2025_GuerillaPlus_Time.xlsx")
excel_sheets("2025_GuerillaPlus_Time.xlsx")
plant_data <- read_excel("2025_GuerillaPlus_Time.xlsx", sheet = "Planting_guerilla")

# if this does not work, just import it via the files tab, right lower corner.

data <- X2025_GuerillaPlus_Time
plant_data <- X2025_GuerillaPlus_Time
```

# Prepare Data frame for analysis

Data looks like 08:24:00, even though its meant like 8 minutes and 42 seconds, so we first need to extract the minutes and seconds, then transfer it to 'total minutes'.

```{r}
unique(data$total_time)

# Tijd als string extraheren
time_str <- format(data$total_time, format = "%H:%M:%S")

# Splits op ":"
time_parts <- strsplit(time_str, ":")

# Corrigeer interpretatie: uren zijn minuten, minuten zijn seconden
data$total_minutes <- sapply(time_parts, function(x) {
  minutes <- as.numeric(x[1])   # stond eerst als 'hours'
  seconds <- as.numeric(x[2])   # stond eerst als 'minutes'
  total <- minutes + seconds / 60
  return(total)
})
```

Now, we calculate the total amount of seconds.

```{r}
# Haal de tijd eruit als HH:MM:SS
time_str <- format(data$total_time, format = "%H:%M:%S")

# Split op ":"
time_parts <- strsplit(time_str, ":")

# Corrigeer interpretatie: HH = minuten, MM = seconden
data$total_seconds <- sapply(time_parts, function(x) {
  minutes <- as.numeric(x[1])   # stond als 'uren'
  seconds <- as.numeric(x[2])   # stond als 'minuten'
  total <- minutes * 60 + seconds
  return(total)
})

```

Next, we want to know how long it takes to prepare **1 TPU**, and **96 TPU's** (the amount for 1 plot).

```{r}
# time needed for 1 TPU
data$minutes_per_TPU <- data$total_minutes / data$amount_tpus
data$seconds_per_TPU <- data$total_seconds / data$amount_tpus

# time needed for 96 TPUs
data$minutes_per_96_TPU <- data$minutes_per_TPU * 96
data$seconds_per_96_TPU <- data$seconds_per_TPU * 96

```

# Now we do the same for the 'planting' time.

```{r}
# check values

head(plant_data$time)

# transform into minutes and seconds

plant_data <- plant_data %>%
  mutate(
    total_seconds = hour(time) * 3600 + minute(time) * 60 + second(time),
    total_minutes = total_seconds / 60
  )
```

We also add experience to our data set. Lea, Stijn and Jim are experienced, Maartje and Alice are not. 1 is experienced, 0 is not experienced.

```{r}
plant_data$experience <- ifelse(plant_data$planter %in% c("Stijn", "Jim", "Lea"), 1, 0)
```

#account for the difference between the different preparation steps. Sorting is for both anchored and unanchored, anchored is only for anchored.

```{r}
# Load prep data
prep_data <- data

prep_data <- prep_data %>%
  mutate(
    task_type = method,  # Save whether it's 'sorting' or 'anchored'
    method = ifelse(method == "sorting", NA, method)  # Keep NA for now
  )

# Separate sorting
sorting_rows <- prep_data %>% filter(task_type == "sorting")

# Duplicate sorting rows for both methods
sorting_for_both <- bind_rows(
  sorting_rows %>% mutate(method = "anchored"),
  sorting_rows %>% mutate(method = "unanchored")
)

# Add back non-sorting prep rows
non_sorting <- prep_data %>% filter(task_type != "sorting")

# Combine all
prep_cleaned <- bind_rows(non_sorting, sorting_for_both) %>%
  mutate(stage = "preparation")

```

# Analysis with average time per method

Average the planting time per location, whcih shows a general difference in time-investment between anchored vs. unanchored.

-   Calculate average preparation time per method and add this to a new dataframe: prep_sumamry

```{r}
prep_summary <- prep_cleaned %>%
  group_by(method) %>%
  summarise(avg_prep = mean(minutes_per_96_TPU, na.rm = TRUE))

```

-   Calculate average planting time per location and method

```{r}
plant_summary <- plant_data %>%
  group_by(method, location) %>%
  summarise(avg_plant = mean(total_minutes))
```

-   Add preparation in our data frame plant_combined

```{r}
plant_combined <- plant_data %>%
  left_join(prep_summary, by = "method")

```

Then, we want to add both planting time (total_minutes) to preparation time (av_prep) total_minutes + avg_prep = Total_time_per_plot

```{r}

plant_combined <- plant_combined %>%
  mutate(total_time_per_plot = total_minutes + avg_prep)

```

-   Run the model

```{r}
model1 <- lmer(total_time_per_plot ~ method + (1 | location), data = plant_combined)
summary(model1)

```

First run first output:

REML criterion at convergence: 140.9

Scaled residuals: Min 1Q Median 3Q Max -1.12698 -0.74917 -0.04525 0.60518 2.50850

Random effects: Groups Name Variance Std.Dev. location (Intercept) 26.71 5.168\
Residual 160.57 12.672\
Number of obs: 19, groups: location, 5

Fixed effects: Estimate Std. Error t value (Intercept) 71.477 5.132 13.929 methodunanchored -32.633 6.059 -5.386

Correlation of Fixed Effects: (Intr) methdnnchrd -0.687

"A GLMM revealed that unanchored transplantation significantly reduced total time per plot compared to anchored methods (estimate = -32.6 minutes, t = -5.39)."

# Visualize this:

```{r}
# Create a data frame with the estimates
plot_data <- data.frame(
  method = c("Anchored", "Unanchored"),
  mean_time = c(71.477, 71.477 - 32.633),
  se = c(5.132, 6.059)
)

# Create the bar plot with error bars
ggplot(plot_data, aes(x = method, y = mean_time, fill = method)) +
  geom_bar(stat = "identity", width = 0.6, color = "black") +
  geom_errorbar(aes(ymin = mean_time - se, ymax = mean_time + se), width = 0.2) +
  labs(
    title = "Time to Prepare and Plant 96 TPUs",
    x = "Method",
    y = "Total Time (minutes)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

# Analysis with experience variation

```{r}
model2 <- lmer(total_time_per_plot ~ method * experience + (1 | location), data = plant_combined)
summary(model2)

```
Now we'll visualize this :)

```{r}
# Get estimated marginal means for all combinations
emm <- emmeans(model2, ~ method * experience)

# Convert to a data frame for plotting
plot_data <- as.data.frame(emm)

# Rename experience for nicer labels
plot_data$experience <- factor(plot_data$experience, labels = c("No experience", "Experienced"))

# Plot
ggplot(plot_data, aes(x = method, y = emmean, fill = experience)) +
  geom_bar(stat = "identity", position = position_dodge(0.7), width = 0.6, color = "black") +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE), 
                position = position_dodge(0.7), width = 0.2) +
  labs(
    title = "Total Time per Method and Experience Level",
    x = "Method",
    y = "Estimated Total Time (minutes)",
    fill = "Experience"
  ) +
  theme_minimal()
```